\section{THE TWELVE HOUSES AND HOUSE RULERSHIP, ADVANCED STUDY}
 

The houses are prediction wise more important than the signs in Vedic Astrology. Most planetary combinations (yogas) are defined in terms of the houses – like the ruler of the second house of livelihood located in the twelfth house of loss being a yoga for poverty. Vedic Astrology regards planets more as house lords than it does as sign rulers. Hence, to proceed with a deeper study of Vedic Astrology, we must examine this issue of house rulership specifically. We should remember the houses that a planet rules from the Ascendant and the Moon. We should look at it first as a house lord and only secondly according to its natural status.

 

While we have already recommended the background of the Houses, much detail will be added here. The house lords will carry the energies of the houses that they rule and influence their condition, their gains and losses as well.

 

For additional information on the houses relative to Raja Yoga, please examine Lesson 4 of the Workbook. For the different house rulerships from different Ascendants and how they function, please examine Lesson 3 of the Workbook (This material helps you understand the signs better as well). This material is extremely important and has a number of charts examined along with it.

 

\subsection{MEANING OF THE HOUSES}

\paragraph{FIRST HOUSE}

Foundation of the chart overall, body, health, vitality, success in life, career, identity, ego, soul, self

\paragraph{SECOND HOUSE}

Livelihood, speech, self-expression, face, early childhood, literary abilities

\paragraph{THIRD HOUSE}

Vitality, prana, brothers and sisters, youth, friends, arms, sense organs (cognitive and motor), skills, arts and crafts, hobbies, interests, information

\paragraph{FOURTH HOUSE}

Mind, emotions, psychology, mother, heart, home, house, vehicle, education (particularly at home), faith, devotion

\paragraph{FIFTH HOUSE}

Intelligence, intellect, education, judgment, values, children, creative intelligence, dharma, past life influences

\paragraph{SIXTH HOUSE}

Enemies, diseases, injuries, distant relatives, foreigners,  litigation, work, struggle, competition

\paragraph{SEVENTH HOUSE}

Partnership, marriage, social projection, career, success, happiness

\paragraph{EIGHTH HOUSE}

Death, longevity, subtle influences, the occult, medicine, Tantra, collective resources, corporations,  loss of fame, deception, calamities

\paragraph{NINTH HOUSE}

Dharma, religion, spirituality, the father, grace, good fortune, luck, higher education, principles, values, belief, conscience

\paragraph{TENTH HOUSE}

Career, vocation, karma, public success, public influence, recognition, fame, elevation

\paragraph{ELEVENTH HOUSE}

Gains, income, achievements, titles, rewards, social recognition, social influence, associates, administrative abilities

\paragraph{TWELFTH HOUSE}

Loss, liberation, debility, selflessness, service, retirement, rest, sleep, samadhi



 

\subsection{HOUSE LORDS}
 

We introduced the basic information on house lords and their meaning in the Astrology of the Seers. Here we will go into more detail on this important and complex subject.

 

Generally, planets are auspicious or inauspicious relative to the houses they rule as calculated from each particular ascendant. This is their “temporal” status, regardess of what their natural status may be.
However, planets are generally good for the affairs of the houses that they rule. Even the lord of a bad house, like the eighth, will be good for the affairs of that house like longevity.

\subsection{HOUSE RULERSHIP TABLE}
TEMPORAL DISPOSITION OF PLANETS
 
\begin{center}
\begin{tabular}{ l l l l l l l l}
Sign &Sun	&Moon	&Mars	&Mer.	&Jup.	&Venus	&Saturn \\
Aries	& 5 A	&4 A	&1, 8 A	&3, 6 I	&9, 12 A	&2, 7 I	&10, 11 I\\
Taurus	&4 N	&3 I	&7, 12 I	&2, 5 A	&8, 11 I	&1, 6 A	&9, 10 *\\
Gem.	&3 I	&2 N&	6, 11 I	&1, 4 A	&7, 10 I	&5, 12 A	&8, 9 N\\
Can.	&2 N	&1 A	&5, 10 *	&3, 12 I	&6, 9 A	&4, 11 I	&7, 8 I\\
Leo	&1 A	1&2 N	&4, 9 *	&2, 11 I	&5, 8 A	&3, 10 I	&6, 7 I\\
Virgo	&12 N	&11 I	&3, 8 I	&1, 10 A	&4, 7 I	&2, 9 A	&5, 6 N\\
Libra	&11 I	&10 N	&2, 7 I	&9, 12 A	&3, 6 I	&1, 8 A	&4, 5 *\\
Scor.	&10 A	&9 A	&1, 6 A	&8, 11 I	&2, 5 A	&7, 12 I	&3, 4 I\\
Sag.	&9 A	&8 N	&5, 12 A	&7, 10 I	&1, 4 A	&6, 11 I	&2, 3 I\\
Cap.	&8 I	&7 N	&4, 11 I	&6, 9 A	&3, 12 I	&5, 10 *	&1, 2 A\\
Aqua.	&7 I	&6 I	&3, 10 I	&5, 8 A	&2 ,11 I	&4, 9 *	&1, 12 A\\
Pisces	&6 N	&5 A	&2, 9 A	&4, 7 I	&1, 10 A	&3, 8 I	&11, 12 I\\
 \end{tabular}
\end{center}
 

A = Auspicious, I = Inauspicious, N = Neutral,

* = Very Auspicious, Raja Yoga Karaka

 

This is a general table and requires more specificity. It should not be applied mechanically:



\subsection{\textbf{Note Audio Explanation of Principles of House Rulership Followed by Detailed Examination}}


The ruler of the ASCENDANT is generally good as it aids in the prospering of the chart overall. Yet if it is a natural malefic or rules another house that is malefic in nature, its benefic status is tainted and it may under certain circumstances give bad results. \\
The ruler of the SECOND HOUSE is generally neutral, though good for wealth and livelihood, which it rules. Yet as a maraka or death-causing planet, it has a negative side as well, particularly when the chart is weak.\\
The ruler of the THIRD HOUSE is generally inauspicious, as it is an impulsive, impetuous force, works through power and can cause oppression or opposition. However, it is usually good for brothers, friends and personal assertion, which the house rules.\\
The ruler of the FOURTH HOUSE, as an angle or kendra, is good if a natural malefic but bad if a natural benefic.  If strong, it is good for the mother and for peace of mind and emotional well-being, can promote devotion.\\
The ruler of the FIFTH HOUSE is generally good, as the fifth is a trine and gives good karmic results, bringing wisdom and good karma into the chart, along with children and creativity.\\
The ruler of the SIXTH HOSUE, a house of disease and injury, is generally bad and can cause harm or at least promote enmity or disease.\\
The ruler of the SEVENTH HOUSE, as an angle, follows the same rules as that of the fourth and is good for the partner. As a maraka, it can cause some trouble for the health later in life.\\
The ruler of the EIGHTH HOUSE, a house of obstacles, opposition and negativity, is generally inauspicious, though it does aid in certain careers like medicine, insurance or corporate positions.\\
The ruler of the NINTH HOUSE, as the best trine and house of fortune, is usually good. It gives grace, good karma, good fortune and well-being, and aids in education and spiritual knowledge.\\
The ruler of the TENTH HOUSE follows the same rules as that of any angular ruler and represents the best angular house. When strong the career and overall success of the person will prosper.\\
The ruler of the ELEVENTH HOSUE is good for income or gains, which its house rules. However, it is malefic for the chart as a whole because it has a disruptive, impulsive, anarchic influence and can cause disease and injury like the sixth lord.\\
The ruler of the TWELFTH HOSUE is generally inauspicious but more often neutral in character, depending upon the other house it rules. It can cause loss but can also connect us to foreign groups or ideas.\\
 

When a planet rules two houses, we must combine the natural status of the planet along with the two houses it rules, as well as its relationship with the ruler of the Ascendant.

 

For example, Saturn rules the eighth and ninth houses for Gemini Ascendant, a good house and a bad house. It is naturally malefic but normally a friend of Mercury, which rules Gemini. Its mulatrikona sign Aquarius governs the ninth. Hence, it would be mixed but slightly good for Gemini. However, if it is a temporal enemy of Mercury in the chart, its results would tend to be bad. Depending upon its aspects and associations, it can function either as the lord of the eighth or ninth houses. Planetary positions in each chart will modify these principles of house rulership.

 

When we classify a planet as neutral or mixed for an ascendant, we mean that it can give both good and bad results, though in different areas of life, not that it will give only neutral effects.

 

When Saturn rules the eighth and ninth, for example, it is still tainted by its eighth house rulership and will give some negative effects. Its effects will not simply cancel each other out but be good for some things and bad for others.



\subsection{PLANETS BY ASCENDANT RELATIVE TO HOUSE RULERSHIP}
 

The following defines each planet by ascendant, its particular affects as lords of certain houses. Never look at a planet just as its natural status like Jupiter as the great benefic Jupiter. Always consider its house rulership relative to the particular ascendant. The indications are typical and can be modified or expanded. This is an important reference table.

 

\subsubsection{Aries Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 5, is a planet of creative intelligence, progeny, counsel and good karma. It is very auspicious. It makes Aries types good lawyers, advisors or psychologists and gives them strong passions and good creativity. Yet it usually indicates that they will not have many children and can burn their hearts out through too much willfulness, impulsiveness, drama and passion.

 

\paragraph{Moon}

The Moon, as ruler of house 4, is a planet of emotion, home, happiness, and the mother. It is generally auspicious but according to some, suffers somewhat by owning an angle as a benefic. Hence, Aries types can suffer from being too emotional or worrying too much. But the Moon is not usually very harmful for them.

 

\paragraph{Mars}

Mars, as ruler of houses 1 and 8, is a planet of the self and the body. It is generally auspicious but tainted by its rulership of the eighth. It is a double indicator of longevity, ruling two houses of longevity. It shows that Aries people may be prone to vice or lack of ethics on one hand, or may conduct some profound research on the other (these being the positive and negative meanings of the eighth house). It shows a potential for them to be headstrong and impulsive, and potentially self‑destructive. Yet it can give deep insight and profound intelligence (the higher aspects of the eighth house).

 

\paragraph{Mercury}

Mercury, as ruler of houses 3 and 6, is a planet of energy, impulse, friendship, injury, disease, egoism, enmity, and prowess. It shows a potential for conflict with friends. Aries types generally possess an overly critical intellect, through which they can become involved in conflict and controversy. They can also suffer from nervous burnout from too much thinking. Hence, Mercury is generally inauspicious for Aries types.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 9 and 12, is a spiritual planet giving grace and showing renunciation through wisdom. Aries types have a strong sense of principle and can be moved by a higher idealism and benevolence. On a more general level, it can give luck, fortune and status. It is very auspicious, perhaps the best planet for Aries. Yet its twelfth house rulership can taint its influence, particularly later on in life.

 

\paragraph{Venus}

Venus, as ruler of houses 2 and 7, is a planet of wealth and relationship, often showing gains through marriage and partnership, or multiple partners. For Aries types, partnership may be bound up with work and finances. They need to work with others to balance out their overly individualistic nature. Venus is inauspicious and becomes a maraka or death-dealing planet when the chart is weak or afflicted..

 

\paragraph{Saturn}

Saturn, as ruler of houses 10 and 11, is a planet of power, prestige, gains, impulse, injury, disease and egoism. Aries types have an ambition for success and achievement that tends to become excessive and must be controlled or it will lead to a fall. They can burn themselves out prematurely through overwork if afflicted. Saturn is perhaps the most inauspicious planet for Aries.

 

\paragraph{Overall}

The Sun and Jupiter (rulers of houses 5 and 9) give spiritual power and character. Mercury and Saturn (ruling houses 6 and 11) show excessive impulsiveness and a critical nature. The Sun and Moon together (ruling houses 4 and 5) give a powerful impact on life and a strong creative nature. The Sun and Moon, or the Moon and Jupiter (lords of angles and rines together) give Raja Yoga (great prestige).

 

Aries is the most one‑sided of ascendants and, usually, the easiest to judge. This is because house and sign rulers always correspond. When the first sign marks the first house, all the other signs will mark the other corresponding houses. Hence, planets are double significators for Aries. The Moon, for example, indicates the mother by its nature and by its rulership of the fourth house. Any influence on the Moon will, therefore, doubly affect the mother. For this reason, Aries ascendant can make for a very strong or a very weak chart.

 

As a cardinal and a fire sign, Aries types act with strength and decisiveness, which may be impulsive or excessive. Their planets manifest this angularity. Aries types are good at starting things but may consume themselves like meteors. They have strong minds and much energy in their heads but can be headstrong.

 

\subsubsection{Taurus Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 4, is a planet of emotion, home and happiness. It shows the tendency of Taurus types towards expansion through property and comfort, as well as their soul tendency towards grace or beauty. Its shows their tendency to worry and possible separation from their home life in their attempt to expand their sphere of material development. It is generally neutral, considered slightly auspicious by some (by being a malefic ruling an angle), slightly inauspicious by others (by its basic nature).

 

\paragraph{Moon}

The Moon, as ruler of house 3, is a planet of energy, interest, motivation, impulse and enthusiasm. It shows the strong sense of curiosity of Taurus types, which may lead them into artistic pursuits on one hand, or superficial hobbies on the other. It is generally inauspicious and shows the tendency of Taurus to emotional impulse and excess.

 

\paragraph{Mars}

Mars, as ruler of houses 7 and 12, is a planet of relationship, sexuality and passion. It indicates an impulse towards sensuality, which may lead to loss of energy and dissipation. It can also show financial loss or expenditure on relationship. It shows the Taurus capacity to aggression in relationship and their high passion levels. It is inauspicious.

 

\paragraph{Mercury}

Mercury, as ruler of houses 2 and 5, is a planet of speech, education, intelligence and creativity. It shows the inherent artistic and poetic sense of Taurus types – their beauty and formal order of expression. They prosper through good communication. It is generally auspicious.

 

\paragraph{Venus}

Venus, as ruler of houses 1 and 6, shows a tendency to work with foreigners. It shows how Taurus types can create their own diseases or enmity in life by their excessive inertia, obstinacy and clinging to form. It is generally auspicious but Venus is tainted somewhat by its rulership of the sixth.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 8 and 11, is a planet of willfulness, violence and destruction. It shows a danger towards excessive material expansion and self‑righteousness, and may cause obesity. It is generally inauspicious and causes excess, but its nature somewhat balanced by Jupiters natural benefic qualities.

 

\paragraph{Saturn}

Saturn, as ruler of houses 9 and 10, gives honor, prestige, power, position and great skill. It shows the enduring drive towards practical achievement in the Taurus nature, which often does not become strong until old age. When Saturn is too strong, Taureans can be cold and controlling, neutralizing their otherwise grace. Saturn is very auspicious giving Raja Yoga.

 

\paragraph{Overall}

The Moon, Mars and Jupiter (rulers of houses 3, 8 and 12) bring danger, rash action, and sudden downfalls. Mercury and Jupiter (ruling houses 5 and 11) together make for profound intelligence. Saturn gives Raja Yoga and great accomplishments, particularly if combined with Mercury, which can give them a spiritual nature. Venus and Mars together make for a highly sensual disposition and possible dissipation. For Taurus Venus gives more success if tempered by Saturn.

 

As a fixed and earth sign, Taurus types are slow to initiate things but very good at perfecting form. They are often conservative and once they are into something are slow to let go. Hence, the planets for Taurus manifest slowly but thoroughly.

 

\subsubsection{Gemini Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 3, shows Geminis soul of curiosity, intellectual impulse and strong opinions, which can cause them to rely too much on their own thoughts and energies. It also shows their strong sense of friendship and tendency to form alliances, particularly for the working out of their ideas. It is inauspicious.

 

\paragraph{Moon}

The Moon, as ruler of house 2, gives fluency of speech, good education and capacity to gain wealth. It endows Geminis with highly developed minds. Additionally, it indicates their great sensitivity and need for emotional nurturing as children. It is generally neutral.

 

\paragraph{Mars}

Mars, as ruler of houses 6 and 11, is triply (by nature and by ruling two houses of injury) a planet of violence. Geminis tend to suffer from accidents and injuries from their absent‑mindedness or through blindness generated by their own opinionatedness or excessive idealism. They suffer from excessive activity, overexertion or hypersensitivity leading to nervous burnout. Mars is very inauspicious for them.

 

\paragraph{Mercury}

Mercury, as ruler of houses 1 and 4, indicates their close alignment of self and mind. While this gives directed mental powers, it can detract from objectivity in regard to the self. It can cause an overly self‑sensitive psychological nature. Any aspects to Mercury will triply affect the mind because the planet rules the self, the ascendant, the mind, and the fourth house and naturally indicates the intellect as well. Mercury is auspicious as the ascendant lord.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 7 and 10, shows a strong will and idealism as well as a seeking of power and prestige in relationship. Geminis over express themselves in relationship, which can lead to dissipation and a diffusion of energy. Though generally inauspicious, it is mainly when weak by house, sign or aspect that Jupiter will give bad results.

 

\paragraph{Venus}

Venus, as ruler of houses 5 and 12, is a planet of passion and luxury. It shows the strong artistic sense and wealth-giving power of Gemini but also its excessive sexuality that leads to loss or dissipation. Gemini types suffer from nervous stimulation and can become enervated through sex. On a higher level, Venus gives them creativity and intelligence and general good fortune. It is probably the most auspicious planet for the ascendant.

 

\paragraph{Saturn}

Saturn, as ruler of houses 8 and 9, can function as a spiritual planet, giving profound intelligence. It also shows the basic ambiguity and contradictions of Gemini types, which can destroy them. They get caught in the dualities of their own minds. Hence, Saturn becomes neutral in its function, though perhaps slightly auspicious (with its Mulatrikona sign Aquarius marking the ninth).

 

\paragraph{Overall}

Venus and Mercury (ruling houses 4 and 5) together give Raja Yoga. Venus and Saturn (rulers of houses 5 and 9) give much success inlife.  Mercury and Saturn (ruling houses 1 and 9) give spiritual awareness. Mars with Jupiter or the Sun (rulers of houses 6, 7 and 11) creates a disturbed or hyperactive mind, which may lead to difficulties or disasters.

 

As a mutable and air sign, planets for Gemini tend to scatter and diffuse their energy. Their action is often quick, premature, irregular and hard to predict. Though a good ascendant for the mind, it is not so good for long term success in the world and has a tendency towards disease and injury. In this computer age, Gemini has a new advantage, however, with their quick and abstract minds.

 

\subsubsection{Cancer Ascendent}
 

\paragraph{Sun}

The Sun, as ruler of house 2, is a planet of speech, wealth, and education, giving good earning capacity and ruling powers. Yet it can give great expenditures and material responsibilities, if afflicted. Cancer types put their soul into their communications. The Sun is generally neutral for them.

 

\paragraph{Moon}

The Moon, as ruler of house 1, is the planet of self and basic orientation in life. It shows that for Cancer types their soul and emotions are closely linked. This affords them a great sympathy and sensitivity that may render them passive or influenced by social trends, which they can also become exponents of. Their heart is in human feeling and emotional considerations are very important for them. The Moon is auspicious as the lord of the ascendant.

 

\paragraph{Mars}

Mars, as ruler of houses 5 and 10, gives great power. Cancer types need to marshal their energies and act with force and discipline in order to succeed in life. Mars is very auspicious and gives Raja Yoga. Yet as a natural malefic can still cause harm at times.

 

\paragraph{Mercury}

Mercury, the ruler of houses 3 and 12, shows their tendency towards loss of energy and mental power through too much emotion. They may also suffer losses through friends, as their alliances may be more based upon sentiment than true affinity or business sense. It is generally inauspicious.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 6 and 9, shows their basic ethical nature and sympathy even towards enemies or foreigners. If weak, it shows losses through them. Excess of emotion over good fortune may cause problems as well. Jupiter is generally auspicious but at times can function negatively as lord of the sixth (particularly when several malefics are in the sixth).

 

\paragraph{Venus}

Venus, as ruler of houses 4 and 11, shows their seeking of property and comfort in the home life. It is their planet of sensuality and shows the sentimentality and attachment they must learn to control. It often gives them gains through property or vehicles. It is generally inauspicious.

 

\paragraph{Saturn}

Saturn, as ruler of houses 7 and 8, shows their tendency to relationships with those of inferior status, who may be hard working or harsh in speech. This may lead them to some kind of disgrace or cause them self-examination. Cancer types tend to suffer in partnership. Their over emotionality may make them manipulative and dominating. Saturn is the most inauspicious planet for them.

 

\paragraph{Overall}

For Cancer ascendant, Mars and Jupiter (rulers of houses 9 and 10) give political and social power and prestige. Mars gives Raja Yoga even by itself. Jupiter and the Sun (ruling houses 2 and 9) give intelligence. Mercury and Venus (ruling houses 3 and 11) give difficulties through too much emotion. Saturn and Venus (ruling 8 and 11) bring difficulties and downfall through excessive materialism.

 

As a cardinal and water sign, Cancer types are able to express their emotions and have an emotional impact on others. Their planets act with gentleness and force combined, seldom in a harsh manner. They often achieve much in life through their ability to relate to others and their diplomatic skills. Their own initial shyness and emotional sensitivity (lunar nature) often slows them down.

 

\subsubsection{Leo Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 1, shows the character, will, ego and soul of Leo types. Leos possess a strong passion for life and self‑expression, a good creative intelligence, as well as a certain aristocracy of the spirit that may degenerate into vanity or ostentatiousness. They can develop charisma but may be self‑dramatic or self‑centered. The Sun is doubly a self‑indicator as both lord and indicator of the first house and so aspects on it will be doubly strong for character issues. It is auspicious as the lord of the ascendant.

 

\paragraph{Moon}

The Moon, as ruler of house 12, is a planet of spirituality, as well as of emotional giving and sympathy. Leos tend to have an underdeveloped social sensitivity, as they have an overdeveloped self‑sense, but once they have identified with someone or something, they are capable of great self‑surrender. On the negative side, this can lead to loss, separation or sorrow. On the positive side, it can give detachment and the capacity for contemplation. It is neutral as the lord of the twelfth and can give bad results if poorly aspected or associated.

 

\paragraph{Mars}

Mars, as ruler of houses 4 and 9, indicates their strong spirit, energy and zeal, through which they can accomplish much in life. Leo types can express the higher qualities of Mars and can develop much perception and insight, through which they can discover an effective mode of action. This can, however, render them overly critical. It is very auspicious giving Raja Yoga.

 

\paragraph{Mercury}

Mercury, as ruler of houses 2 and 11, shows their capacity for great material and work gains, high income, with Mercury a double significator of wealth. Leos can usually do well in the business world, as they know how to project themselves, but they can also talk too much about themselves. It is generally inauspicious for the other domains of life.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 5 and 8, is a planet of profound intelligence and deep research. Yet it can be a planet of passion and vice, showing difficulties or dishonor arising from passion. It can also show difficulties with children. Leo types must learn to employ their creativity on the deeper levels of the mind or it can overwhelm them as passion arising from the subconscious. They can suffer from being overly dramatic or overly romantic. It is generally auspicious but tainted by its rulership of the eighth.

 

\paragraph{Venus}

Venus, as ruler of houses 3 and 10, is a planet of power, prestige and egoism. It shows the danger of excessive pride, vanity and ambition. Yet it also shows a capacity for gaining friendship or working with people in high places. It is generally impulsive and inauspicious.

 

\paragraph{Saturn}

Saturn, as ruler of houses 6 and 7, is a planet of enmity and disease, as well as relationship. Leos often tend to subordinate their partners or choose one lower in status in life. Partners may become servants or, revolting from the dominant Leo personality, may become enemies. It is the most inauspicious of the planets for Leo.

 

\paragraph{Overall}

Mercury and Mars (ruling four houses of wealth) together for Leo ascendant give great uwealth and luck. Strong Mars by itself gives Raja Yoga. Mars and Jupiter (rulers of houses 5 and 9) give good karma or spiritual growth, particularly if aligned with the Moon (lord of house 12). Venus and Saturn (rulers of houses 3 and 6) give vanity and manipulation of others.

 

Leo as fixed fire shows a great deal of strength of will and determination. Its planets thus tend to act with both force and consistency. Leos have a strong sense of self and are usually able to achieve what they focus on. On the other hand, if their Sun and sense of self is weak, they can be among the weakest and most pitiable of people. And when strong Leo types fall, they can fall hard.

 

\subsubsection{Virgo Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 12, shows the tendency of Virgo types towards selfless service and self‑abnegation. On a lower level, this can indicate servility, lack of self‑esteem or self‑worth. On a higher level, it gives a great capacity for spiritual work or self‑surrender to any difficult labor. This capacity to negate or modify the self can also give them talents for acting, or psychological difficulties from not having any strong self‑identity. It is generally neutral and benefits from association with a strong Mercury, but its basic malefic nature often comes out.

 

\paragraph{Moon}

The Moon, as ruler of house 11, shows their capacities for great gains through mental work or social communication. They can easily work in a group or with the public as it is hard for them to stand on their own. It is generally impulsive and inauspicious.

 

\paragraph{Mercury}

Mercury, as ruler of houses 1 and 10, is a planet of both self and career. Virgos tend to identify strongly with their work and vocation. They are their work and their work is their life. It is auspicious as the lord of the ascendant.

 

\paragraph{Mars}

Mars, as ruler of houses 3 and 8, is doubly a planet of death and longevity. It shows their tendency to suffer from accidents, injuries, nervous burn out or neuromuscular disorders. Mercury types must be careful not to overly expose themselves to danger and risks, as they suffer through recklessness, and also may be attacked or harmed by others. Their mental energy is much higher than their physical and they can push their physical too far unknowingly. It is very inauspicious and often causes much harm and misery.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 4 and 7, is a planet of emotion, home and relationship. Virgos require a strong basis in home and relationship for the development of their energy in life without which there is lack of self‑confidence. Though generally inauspicious, it is mainly when weak that it causes problems.

 

\paragraph{Venus}

Venus, as ruler of houses 2 and 9, is a planet of grace, luck, fortune and success. Virgo types prosper through artistic expression and through ventures involving teaching and communication. Spiritually, they grow through the development of discrimination in the philosophical mind. It is usually the best planet for Virgo.

 

\paragraph{Saturn}

Saturn, as ruler of houses 5 and 6, shows their tendency to overwork, which can lead to disease. It shows they tend not to have so much happiness through children and can be blocked in their emotional or romantic expression. Health or disease for them is a matter of balancing work with play and exercise. It is of a neutral or mixed character, though it often causes much health problems, particularly when allied with Mars.

 

\paragraph{Overall}

Venus along with Mercury (rulers of houses 9 and 10) can give Raja Yoga, great prestige. Venus with Saturn (ruling houses 5 and 9) gives good intelligence and spiritual development. Mars with Saturn (rulers of houses 6 and 8) give chronic diseases, often of a nervous nature. The Moon with Venus (rulers of houses 9 and 11) can give great wealth.

 

As a mutable and earth sign, Virgo demonstrates a great deal of sensitivity and adaptability on a physical level, as well as mental power and expressive skills. Its planets function with subtlety and variability. Yet there is a tendency towards weakness or dissipation, debility or disease.

 

\subsubsection{Libra Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 11, is a planet of gains, aspirations and ideals, yet also impulse, power, prowess, violence, disease and enmity.  Librans tend to have an idealistic egoism and self‑righteousness. They have great goals for themselves that are hard to achieve. They can attempt too much and overextend themselves. The Sun is generally impulsive and inauspicious for Libra.

 

\paragraph{Moon}

The Moon, as ruler of house 10, is a planet of power, prestige, action and realization. Librans gain influence and fame through a lunar sympathy with others. On a political level, this can give them power over the masses. They are usually progressive and aware of social trends, which they help initiate. The Moon is generally neutral but sometimes considered inauspicious (as a benefic ruling an angle).

 

\paragraph{Mars}

Mars, as ruler of houses 2 and 7, is a planet of wealth and relationship and can indicate conflict over finances. Mars ruling the seventh makes relationship a potential source of conflict. Libras can project an aggressive energy into relationship or experience one through it.  Mars is generally inauspicious for Libras.

 

\paragraph{Mercury}

Mercury, as ruler of houses 9 and 12, is a spiritual planet giving renunciation through wisdom, an idealistic sense of self‑sacrifice. Libran spirituality requires a strong intellectual base (Mercury) as they tend to be carried away by speculation and ambition (Jupiter). Mercury is generally the best planet for Libra both for the affairs of the inner as well as the outer life.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 3 and 6, is a planet of energy, impulse, gains, ambition, egoism, disease and enmity. It shows conflict with friends, a critical mind, and a tendency to political or religious zeal. Librans suffer from trying too much to change the world. Though generally inauspicious, Jupiter is not always so and can give good results if it is aspected by Mercury and Venus.

 

\paragraph{Venus}

Venus, as ruler of houses 1 and 8, represents self, life and longevity. It shows a tendency towards vice or passion on one hand, or profound idealism and aspiration on the other. Though generally auspicious as the ruler of the ascendant, it is tainted by its rulership of the eighth.

 

\paragraph{Saturn}

Saturn, as ruler of houses 4 and 5, is a planet of intelligence and happiness. It indicates Librans tenacity, concentration, endurance and profundity, through which they succeed in life. Librans are stern about their principles and, when these are pure, can accomplish much inwardly as well as outwardly. Saturn is very auspicious, giving Raja Yoga.

 

\paragraph{Overall}

Saturn and Mercury (lords of houses 5 and 9) in combination give spiritual growth. Saturn by itself (lord of houses 4 and 5) gives Raja Yoga or with the Moon (ruling house 10). Jupiter with Mercury and Venus gives a high idealism and love of truth and harmony. Jupiter with Mars can create violence, egoism or disease.

 

As a cardinal and air sign, Librans act with decisiveness once they are clear about their principles. Libra is often considered to be the best Ascendant, as the main negative planet becomes Jupiter, whose great benefic nature often makes it function in a positive way, even against the rules of house lordship.

 

\subsubsection{Scorpio Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 10, is a planet of power, prestige and position and indicates the nature of the individuals life action and realization. Scorpio types possess a strong will to succeed and to prevail and will work behind the scenes or in a devious manner to achieve their ends. The Sun is considered generally auspicious (as a malefic ruling an angle).

 

\paragraph{Moon}

The Moon, as ruler of house 9, is a spiritual planet indicating dharma, the guiding principle in life and the grace or fortune that follows it. Spirituality for Scorpio types is in lunar qualities of receptivity, devotion and contemplation. The Moon is perhaps their best planet, ruling the best of trines.

 

\paragraph{Mars}

Mars, as ruler of houses 1 and 6, represents self and disease, both prosperity and difficulty. Scorpios tend to cause their own diseases or become their own worst enemies, as they tend to be too willful, fixed and obstinate in their behavior. Mars, though generally auspicious as the ruler of the ascendant, is tainted by its rulership of the sixth house.

 

\paragraph{Mercury}

Mercury, as ruler of houses 8 and 11, is a planet of destruction, violence, impulse and aggression, of injury arising from subconscious emotions. Yet it also shows the possibility of gain through profound thinking. Scorpios strong critical intelligence is a two-edged sword and can harm the one who wields it. Though generally inauspicious, Mercury cannot cause as much harm for Mars-ruled ascendants as Mars does since its basic nature is benefic.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 2 and 5, is a planet of speech, intelligence, education and wealth. It shows gain through children. Scorpios are often expressive, articulate and have good creative faculties. They may be good in sports or athletics, as they often possess good muscles. For them, Jupiter is very auspicious.

 

\paragraph{Venus}

Venus, as ruler of houses 7 and 12, is a planet of passion and shows the potential of loss and sorrow through the marriage partner. Scorpio types tend towards excessive sexual passion, luxury and extravagance with such a sensual Venus. Venus is thus an inauspicious planet for this Ascendant.

 

\paragraph{Saturn}

Saturn, as ruler of houses 3 and 4, shows strong energy and sometimes violence arising from excessive impulse in the emotional nature. There is a strong drive to acquire property, to participate in public affairs and to benefit from friends. Saturn is probably the worst planet for Scorpio.

 

\paragraph{Overall}

The Sun with the Moon (rulers of houses 9 and 10) give Raja Yoga, status and prestige. Jupiter (ruler of house 5) also gives general prosperity. Jupiter and the Moon (rulers of 5 and 9) give spiritual knowledge and a Sattvic (pure) mind. Mars with Mercury or Saturn (rulers of houses 6, 8 and 3) gives an aggressive disposition with hostility or inflicting of injury.

 

As a fixed and water sign, Scorpios are stable and determined and do not change quickly. Their planetary influences tend to remain constant and often get stuck as it were. They often work behind the scenes and are hard to see for what they really are.

 

\subsubsection{Sagittarius Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 9, shows the basic ethical or spiritual orientation of the Sagittarius soul, its sense of law, harmony and beneficence, its inherent capacity to rule and give favor. Yet, on the negative side, it can show a too egoistical sense of principle. It is very auspicious.

 

\paragraph{Moon}

The Moon, as ruler of house 8, shows their profound emotionality that gives them an interest in what transcends death. Their dedication to their principles is often to the death; if highly afflicted, they may kill others for them. It is generally neutral but can cause problems, like disease, as the lord of the eighth, if under malefic influences.

 

\paragraph{Mars}

Mars, as ruler of houses 5 and 12, shows their strong passion and creativity which, if afflicted, can indicate loss through passion or progeny. On the positive side, it gives good perception and possible spiritual discrimination. It gives gains and is generally auspicious.

 

\paragraph{Mercury}

Mercury, as ruler of houses 7 and 10, shows their basic mental impulse toward power, prestige and social preeminence that can lead them into excessive action. Sagittarius types may seek partnership based upon status and such relationships may be transient or superficial. Mercury is generally inauspicious, particularly when weak in the chart.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 1 and 4, shows their close identification of self and mind. They can be influenced by traditional or family beliefs and possess much faith and contentment in life. This can make them too conservative, however. Jupiter is auspicious as the lord of the ascendant.

 

\paragraph{Venus}

Venus, as ruler of houses 6 and 11, shows a tendency towards excess enjoyment, self‑indulgence and extravagance. Sagittarius types may be overly confident and optimistic and lack the capacity to discern between their own wishes and the good of all. They suffer from indulging in their beneficence, just as they prosper in purifying it. Venus is rather inauspicious for them.

 

\paragraph{Saturn}

Saturn, as ruler of houses 2 and 3, shows an impulse towards wealth that can be manipulative. There may be a harshness of speech that can cause enmity. These types are strict about financial security, though they can be careless with what they think is extra. Saturn is probably their worst planet but Venus is nearly as bad.

 

\paragraph{Overall}

The Sun and Mars can give Raja Yoga, great prestige, and are generally good for prosperity and beneficence. The Sun and Jupiter (ruling houses 4 and 9) give spiritual knowledge. Venus and Saturn (rulers of houses 3 and 6) together give vanity and a materialistic outlook on life, which can bring disease. Saturn can cause a downfall and warp their values.

 

As a mutable and fire sign, Sagittarius types have the most critical minds. They are dramatic and expressive but not always prudent. Their planets tend act quickly and decisi­vely but not always consistently.

 

\subsubsection{Capricorn Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 8, shows the tendency of Capricorn types towards excessive willfulness. They have a dark sense of themselves that can manifest as destructive ego or on a higher level, as the destruction of the ego. This can manifest as a selfish materialistic attitude in life. On a psychic level, it can create a black magician. It can also give the capacity for profound research. Once this will is mastered, it can create the strongest of spiritual aspirations, the will to negate the ego. The Sun is very inauspicious for them.

 

\paragraph{Moon}

The Moon, as ruler of house 7, indicates their strong need for relationship to balance out their generally cold disposi­tion. It shows their strong social sense, which is not really sensitive but rather detached, controlling or political. The Moon is generally neutral but some consider it inauspicious (as a benefic ruling an angle).

 

\paragraph{Mars}

Mars, as ruler of houses 4 and 11, shows their desire for property and income. It also gives them a harsh or penetrating power of mind that can, if afflicted, lead to violence. It is inauspicious.

 

\paragraph{Mercury}

Mercury, as ruler of houses 6 and 9, shows their ambiva­lence towards principles and ideals, which can lead to opiniona­ted­ness or conflict. On the positive side, they can be extremely hard workers and give dedicated service to manifest materially what they really believe in. It is generally auspicious but can function negatively if the lord of the sixth is afflicted.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 3 and 12, shows the Capricor­nian tendency towards loss or sorrow through too much greed and manipulative attitudes. They can suffer through friends or alliances and are often forced to make it on their own. Jupiter is generally inauspicious for them.

 

\paragraph{Venus}

Venus, as ruler of houses 5 and 10, gives Capricorn a great power of success through an adaptation of grace to their strong but often unrefined nature. On a lower level, this may be utilitarianism, on a higher level, simple fitness of form where content shines unhindered by glamour. Venus is very auspicious for them and gives Raja Yoga.

 

\paragraph{Saturn}

Saturn, as ruler of houses 1 and 2, shows how intimately their self‑sense is bound up with work and livelihood, They are often harsh in speech and forced to work hard in life. Yet they are also self‑disciplined and through their persistence can make great and enduring accomplishments in the long run. Saturn is generally auspicious as the lord of the ascendant.

 

\paragraph{Overall}

Mars and Saturn (malefic lords of houses 1 and 4) combined can give an aggressive mind that may be violent and seek power at all costs. The Sun may work in this way like Mars. Venus by itself gives Raja Yoga or great prestige. Venus with Mercury (rulers of houses 9 and 10) give Raja Yoga as well as spiritual development. Saturn and Mercury (rulers of houses 1 and 9) enhance spirituality.

 

As a cardinal and earth sign, Capricorns act with force and substance. Their planets act slowly but strongly and often with an obstinate determination. They may start slow in life or at a low level. They usually win out in the end through a long-term plodding action.

 

\subsubsection{Aquarius Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 7, shows the tendency of Aquarius types to become dominated by others and to seek refuge through partnership or relationship. They usually marry superi­ors or take an inferior role towards their partners, or at least a role of service. The Sun is generally inauspicious here (though some consider that it benefits by being a malefic lord of an angle).

 

\paragraph{Moon}

The Moon, as ruler of house 6, shows their strong capacity to put their minds and emotions into work and service. It also shows their tendency towards disease from overly sensitive feelings and too much mental activity. It is generally inauspicious and often causes health problems.

 

\paragraph{Mars}

Mars, as ruler of houses 3 and 10, shows their strong energy towards authorities and institutions. Sometimes they work for them as selfless workers. Other times they work against them as revolutionaries. It is generally impulsive and inauspicious.

 

\paragraph{Mercury}

Mercury, as ruler of houses 5 and 8, shows their capacity for profound intelligence, deep and philosophical thinking and understanding the mysteries of the psyche. Their minds are profound, ambiguous or dark. They may also suffer through children. It is generally auspicious but tainted by its rulership of the eighth (it is usually less auspicious than it is for Capricorn).

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 2 and 11, is a great planet of wealth and income. Aquarius types have a great capacity for material gain through their work once they get beyond their self‑subordination in life as they are capable of saving their resources until they can accumulate to the point of being really worth something. They will slave away for what they believe in and this may give them great rewards in time. Jupiter is a little inauspicious by house rulership.

 

\paragraph{Venus}

Venus, as ruler of houses 4 and 9, shows their basic sensitive and spiritual nature. Through it, they can gain power and success in life by manifesting the ethical ideals of their faith or belief. On a lower level, however, this may just be social pretence and adaptation to a group that may be devoid of ethics. Venus is a very auspicious planet and gives Raja Yoga.

 

\paragraph{Saturn}

Saturn, as ruler of houses 1 and 12, shows the Aquarian tendency towards self‑negation. In a higher sense, this can be selfless service or the negation of the ego into the divine. On a lower level, it can be servility and self‑degradation, the abuse of self. Whatever they have faith in they are willing to sacrifice all for, not only for themselves, but also for others. If it is a faith in God, it is greatly exalting. If it is a faith in a human group, like the mafia, it is greatly destruc­tive. Saturn is generally auspicious as the lord of the ascendant but as the ruler of the twelfth it becomes perhaps the weakest of the ascendant lords.

 

\paragraph{Overall}

Venus produces Raja Yoga, particularly when combined with Mars (ruler of house 10). Yet this tends to be a rather worldly combination. Saturn and Venus (rulers of houses 1 and 9) give more spirituality or integrity. Mercury and Jupiter together (rulers of 2 and 5) give good intelligence or good income.

 

Aquarius, as a fixed and air sign, shows a consistency of belief that is often unaware of practical realities. Its planets act with mystery and ambiguity, or in a subterranean manner. Aquarian concerns are usually not with the visible world but with the matters of faith and association.

 

\subsubsection{Pisces Ascendant}
 

\paragraph{Sun}

The Sun, as ruler of house 6, shows the tendency of Pisces types to suffer from enemies, diseases or opposition. This causes them either to become very strong in themselves, some­times even hostile and aggressive, or to take a background role in life and drift irresolutely. The Sun for them is generally inauspicious.

 

\paragraph{Moon}

The Moon, as ruler of house 5, shows their strong creative intelligence, their love of imagination and speculation, but also their tendency to be overly emotional or sentimental. It gives a love of children. The Moon is very auspicious for them as the lord of the house of good karma.

 

\paragraph{Mars}

Mars, as ruler of houses 2 and 9, is their planet of grace and fortune. Pisces types prosper through learning to logically express and independently manifest their higher intuition. They benefit through the energy and decisive action of Mars. Mars is very auspicious for them.

 

\paragraph{Mercury}

Mercury, as ruler of houses 4 and 7, is a planet of home, happiness and relationship. The Pisces mind has a strong domestic orientation. They may marry at a young age and remarry later in a more stable way. Relation­ships may be superficial or fluctuating. They may be overly sensitive to them and hence prone to more worry about home and partner. Mercury is generally inauspicious, particularly when weak.

 

\paragraph{Jupiter}

Jupiter, as ruler of houses 1 and 10, shows their strong need for outer success and social recognition in life to prove their self‑worth. It can render them somewhat conservative or identified with the government or country, followers of tradi­tion and authority, or very orthodox in their beliefs. Jupiter is benefic as the lord of the ascendant.

 

\paragraph{Venus}

Venus, as ruler of houses 3 and 8, is their indicator of disease and misfortune. Pisces types suffer from self‑indul­gence, dissipation, and excess consumption of sugar, alcoholism, drugs and other forms of escapism. They may also come under the psychic or commercial domination of other people. Venus is rather inauspicious for them but largely because of their own actions.

 

\paragraph{Saturn}

Saturn, as ruler of houses 11 and 12, shows their alternat­ing fortunes, their fluctuations between gain and loss, increase and decrease. Saturn here is a powerful malefic and can cause disease or suffering through a Piscean over sensitivity to the fluctuations of life.

 

\paragraph{Overall}

Mars with Jupiter (rulers of 9 and 10) or the Moon with Jupiter (rulers of 5 and 10) can create Raja Yoga. Venus and Saturn (rulers of 8 and 11) can produce disease and misfortune. Mars with the Moon (rulers of 5 and 9) give good intelligence and beneficence. The Moon with Venus (rulers of 5 and 8) can give a profound or troubled intelligence.

 

Pisces, as a mutable and water sign, shows emotional sensitivity and variability. Its planets act in a similar dualistic and changeable manner. When their enthusiasm is high, they can accomplish much, but they find it hard to sustain a consistency of action. Pisces, in spite of being ruled by Jupiter, can be a difficult ascendant, particularly for outer success in life. The psychology is also either very intuitive or highly vulnerable.

 

\subsubsection{Summary}

In summary, it is important to note the specific energetics of each Ascendant. Planetary influences may combine together by aspect, by being located in the sign ruled by another planet, or by ruling or residing in related houses (like two houses of wealth, the second and eleventh).

 

Try to see the potential high and potential low of each Ascendant and the role each planet is likely to play for it. Note not only house rulership but also the influences planets may project by aspect and association.